Xolotl: An Intuitive and Approachable Neuron \& Network Simulator in MATLAB

### 2018-07-10 paper/paper.tex
wrote caption for mechanism figure

	Conductance-based models of neuron dynamics based on Hodgkin-Huxley formalism have been used extensively in computational neuroscience. However, working with these models continues to be a challenge due to their high dimensionality and large number of parameters, hindering intuition of their dynamical features. Here, we present a neuron and network simulator built on a novel automatic type system that binds class templates written in "C++" to object-oriented code in "MATLAB". Our approach builds on the tradition of uniting the speed and power of languages like "C++" with the ease-of-use and rich libraries of scientific programming languages like "MATLAB". In our software, models are hierarchical, with components that are named and searchable, permitting intuitive high-level programmatic control over all parameters. Our simulator's architecture allows for the live manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. Our simulator is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary mechanisms and dynamics. Finally, the simulator is written in a modular fashion and has been released under a permissive free software license, enabling it to be integrated easily into the workflow of many researchers.

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_HH}
	\caption{\texttt{xolotl} can quickly set up and simulate conductance-based models. Schematic representation of a electrotonically-compact neuron with three populations of ion channels (colored rectangles). In the simulator, the soma is represented using an object of class $compartment$ and the populations of ion channels are represented by $conductance$ objects contained within the compartment object. Code snippet used to set up this neuron model, inject current, integrate and plot the voltage, and show activation functions demonstrates how various several steps can be run in a few lines of code. (A) Simulated voltage trace of a Hodgkin-Huxley model with three conductances and 0.2 nA of injected current. Colors indicate the dominant current (gold is fast sodium (NaV), blue is delayed rectifier (Kd), red is Leak). (C) f-I curve of this neuron. (D-E) Steady-state gating functions for activation (m) and inactivation (h) gating variables. (F-I) Voltage-dependence of time constants for activation (m) and inactivation (h) gating variables}
	\label{fig:figurehh}
\end{figure}

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_clamp}
	\caption{Simulating a voltage-clamp experiment. The cartoon shows a cell with delayed rectifier potassium conductance (\cite{liuModelNeuronActivitydependent1998}) that is being recorded from in two-electrode voltage clamp. The code snippet sets up a model with a single compartment and a single channel type, and clamps the cell to a constant voltage and integrates it. (A) Voltage steps that the cell is clamped to. (B) Current injected into the cell to clamp it to the voltages in (A). (C) Asymptotic clamped current vs. clamped voltage for this cell. (D) Accounting for the reversal potential of Potassium ions yeilds the conductance-voltage curve of this channel type. (E) Normalized conductance-voltage curves, with sigmoid fits with various exponents. (F) An exponent of $n=4$ yields the best fit, allowing for the characterization of the activation function of this channel type. }
	\label{fig:figureclamp}
\end{figure}

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_mechanism}
	\caption{Snapshots allow the user to bookmark points in parameter and state space of the model and return to them when needed. The initial state (orange disc) of the single compartment model in the previous example is saved using the snapshot method. This saves all parameters and dynamic variables of the model in a named state. The first column shows the profile of conductances and the voltage dynamics of the model at this point in time. First, the maximal conductances of the Calcium channels are set to zero (purple), changing the voltage dynamics of the neuron. After evolving the model for some time (green), the conductance profile and voltage dynamics returns to a state similar to the initial state. This configuration is now saved in a state called final and the initial configuration is returned to using the \texttt{reset} method (backwards arrow from green to orange).  This configuration is now saved in a state called final and the initial configuration is returned to using the \texttt{reset} method (backwards arrow from green to orange). In summary, any state can be bookmarked using a descriptive name using the \texttt{snapshot} method, and can be returned to using the \texttt{reset} method.}
	\label{fig:figuremechanism}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		INTEGRAL CONTROL
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Simulating Homeostasis}

>>>>

	Conductance-based models of neuron dynamics based on Hodgkin-Huxley formalism have been used extensively in computational neuroscience. However, working with these models continues to be a challenge due to their high dimensionality and large number of parameters, hindering intuition of their dynamical features. Here, we present a neuron and network simulator built on a novel automatic type system that binds class templates written in "C++" to object-oriented code in "MATLAB". Our approach builds on the tradition of uniting the speed and power of languages like "C++" with the ease-of-use and rich libraries of scientific programming languages like "MATLAB". Our software allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parameters. Our simulator's architecture allows for the live manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. Our simulator is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary mechanisms and dynamics. Finally, the simulator is written in a modular fashion and has been released under a permissive free software license, enabling it to be integrated easily into the workflow of many researchers.

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_HH}
	\caption{\texttt{xolotl} can quickly set up and simulate conductance-based models. Schematic representation of a electrotonically-compact neuron with three populations of ion channels (colored rectangles). In the simulator, the soma is represented using an object of class $compartment$ and the populations of ion channels are represented by $conductance$ objects contained within the compartment object. Code snippet used to set up this neuron model, inject current, integrate and plot the voltage, and show activation functions demonstrates how several steps can be run in a few lines of code. (A) Simulated voltage trace of a Hodgkin-Huxley model with three conductances and 0.2 nA of injected current. Colors indicate the dominant current (gold is fast sodium (NaV), blue is delayed rectifier (Kd), red is Leak). (C) f-I curve of this neuron. (D-E) Steady-state gating functions for activation (m) and inactivation (h) gating variables. (F-I) Voltage-dependence of time constants for activation (m) and inactivation (h) gating variables}
	\label{fig:figurehh}
\end{figure}

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_clamp}
	\caption{Simulating a voltage-clamp experiment. The diagram shows a cell with delayed rectifier potassium conductance (\cite{liuModelNeuronActivitydependent1998}) that is being recorded from in two-electrode voltage clamp. The code snippet shown here sets up a model with a single compartment and a single channel type, and clamps the cell to a constant voltage and integrates it. (A) Voltage steps that the cell is clamped to. (B) Current injected into the cell to clamp it to the voltages in (A). (C) Asymptotic clamped current vs. clamped voltage for this cell. (D) Accounting for the reversal potential of Potassium ions yields the conductance-voltage curve of this channel type. (E) Normalized conductance-voltage curves, with sigmoid fits with various exponents. (F) An exponent of $n=4$ yields the best fit, allowing for the characterization of the activation function of this channel type. }
	\label{fig:figureclamp}
\end{figure}

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_mechanism}
	\caption{Adding arbitrary intracellular mechanisms. (A) A single-compartment neuron model with 8 channel types. With these parameters, this model neuron is a tonic spiker. Since there is no mechanism for changing intracellular Calcium in this model, the Calcium level stays constant (top panel). (B)  An intracellular buffering of Calcium and the effect of Calcium influx through voltage gated Calcium channels (VGCCs) is modeled using a simple differential equation. Adding this mechanism causes the cell to burst periodically, with synchronized oscillations in intracellular Calcium. (C) Schematic of Calcium-dependent intregral control homeostastis. In this feedback system, the rates of production of mRNA depends on the Calcium levels in the cell, which depends on the membrane voltage, which in turn depends on the conductance density of all channel types, which is a function of the mRNA abundance. These integral controllers are implemented as a \texttt{mechanism} object, and can be added to conductances. On integrating the model, intracellular calcium levels rise and approach the target (red dashed line). This is accompanied by an increase in the conductance densities of all channels being controlled by this homeostatic mechanism. The voltage behavior of the cell changes from silence to bursting with truncated spikes to regular bursting.}
	\label{fig:figuremechanism}
\end{figure}

@@
